Performing and evaluating graph-based clustering with rOpenScPCA

### 2024-11-08 analyses/hello-clusters/01_perform-evaluate-clustering.Rmd
Apply suggestions from code review
Co-authored-by: Joshua Shapiro <josh.shapiro@ccdatalab.org>
WIP: began sketching out notebook to eval clustering

---
title: "Performing and evaluating graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

@@

@@

```{r packages}
# load required packages
library(rOpenScPCA)
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
})
```

### Paths

Setting paths to directories and files for input and output at the start of a notebook can be helpful for organization and reproducibility.
You may not know all of the paths you need when you start a notebook, but once you do, keeping them all in one place is helpful, both for yourself in the future and for other users of the notebook.

#### Base directories

In the chunk below, we include some common directories that you might need, finding their locations using the very useful [`rprojroot` package](https://rprojroot.r-lib.org/).

```{r base paths}
# The base path for the OpenScPCA repository, found by its (hidden) .git directory
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

@@

# The path to this module
module_base <- file.path(repository_base, "analyses", "hell-c")
```

@@

```{r paths}
# Input files

# Output files
```

## Analysis content

Organize the remainder of your content into sections and subsections as appropriate for your analysis.

Add a new code chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I* on Windows or Linux, *Cmd+Option+I* on a Mac.
Be sure to comment your code and explain your steps as you go.

When you save the notebook, an HTML file containing the code and output will be saved alongside it.
To preview the HTML file, click the *Preview* button or press *Ctrl+Shift+K* on Windows or Linux, *Cmd+Shift+K* on a Mac.

The preview shows you a rendered HTML copy of the contents of the editor.
Consequently, *Preview* does not run any R code chunks.
Instead, the output of the chunk when it was last run in the editor is displayed.

It is usually a good idea to rerun the entire notebook before sharing any output, to ensure that the notebook and output files are fully up to date.
You can do this by clicking on the *Run* menu in the toolbar and clicking on *Restart R and Run All Chunks*.

>>>>

---
title: "Performing and evaluating graph-based clustering with rOpenScPCA"
date: "`r Sys.Date()`"
author: Data Lab
params:
  seed: 2024
output: 
  html_notebook: 
    toc: yes
    toc_float: yes
    df_print: paged
---

@@

This notebook provides examples of how to use the `rOpenScPCA` package to:

- Perform clustering on an SCE object or Seurat object
- Add clustering results to the object
- Evaluate clustering results using a variety of metrics, including:
  - Average silhouette width
  - Neighborhood purity
  - Cluster stability, as measured with the adjusted Rand index

This notebook will use the `SCPCS000001` sample for demonstration and assumes this sample is present in the `OpenScPCA-analysis/data/current` directory.
Please [see this documentation](https://openscpca.readthedocs.io/en/latest/getting-started/accessing-resources/getting-access-to-data/) for more information about downloading OpenScPCA data.

@@

```{r packages}
library(rOpenScPCA)

suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(Seurat)
  library(dplyr)
  library(ggplot2)
})

# Set ggplot theme
theme_set(theme_bw())
```

#### Paths

```{r base paths}
# The base path for the OpenScPCA repository
repository_base <- rprojroot::find_root(rprojroot::is_git_root)

@@

# The path to this module
module_base <- file.path(repository_base, "analyses", "hello-clusters")
```

@@

```{r paths}
# Path to processed SCE file for sample SCPCS000001
input_sce_file <- file.path(data_dir, "SCPCP000001", "SCPCS000001", "SCPCL000001_processed.rds")

# Path to output file where clustering results will be saved
cluster_tsv <- file.path(module_base, "results", "SCPCS000001-cluster-results.tsv")
```

### Read in data

```{r}
sce <- readRDS(input_sce_file)

# Convert to Seurat object for demonstration
seurat_obj <- as.Seurat(
  sce,
  counts = "counts",
  data = "logcounts"
)
```

## Perform clustering

This section will show how to perform clustering with the function `rOpenScPCA::calculate_clusters`.

By default, this function will use the following parameters:

- Louvain algorithm
- Jaccard weighting
- 10 nearest neighbors
- A resolution parameter of 1

These parameters can be customized, as follows:

- `algorithm` can be one of
  - `"louvain"`, `"walktrap"`, or `"leiden"`
- etc.

This function will return a table with both cluster identities and specified parameter values.

## Cluster on an SCE

Make sure to provide the seed!

The function assumes the PC name is `"PCA"`. If it's not, you can provide this ifnormation with the `pc_name` argument.

```{r}
cluster_results <- calculate_clusters(
  sce,
  seed = params$seed
)
cluster_results
```

If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.

_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
You may therefore want to create a new column, for example `openscpca_cluster`

```{r}
# First, extract the colData into a separate data frame for manipulation
sce_coldata <- colData(sce) |>
  as.data.frame()

# Second, let's rename the cluster column in `cluster_results` to avoid conflict
# But, we don't want to overwrite the results themselves
cluster_results_renamed <- cluster_results |>
  dplyr::rename(openscpca_cluster = cluster)

# Third, add columns as desired using the barcodes and cell_id columns to join
sce_coldata <- sce_coldata |>
  dplyr::left_join(
    cluster_results_renamed,
    by = c("barcodes" = "cell_id")
  )

# Finally, add the modified colData back into the SCE, ensuring the rownames are retained
colData(sce) <- DataFrame(
  sce_coldata,
  row.names = sce_coldata$barcodes
)

colData(sce)
```

## Cluster on a Seurat object

```{r}
cluster_results <- calculate_clusters(
  seurat_obj,
  seed = params$seed,
  # Since we converted this from an SCE object, its PC name is `"PCA"`
  pc_name = "PCA"
)

cluster_results
```

If you want to include these in your SCE object colData, you can add columns as shown in the following chunk.

_Caution_! There already exists a column `cluster` which represents an automated clustering which was _not evaluated_ or parameterized according to the dataset. 
You may therefore want to create a new column, for example `openscpca_cluster`

```{r}
# Let's rename the cluster column in `cluster_results` to avoid conflict and/or ambiguity
cluster_results <- cluster_results |>
  dplyr::rename(openscpca_cluster = cluster)

seurat_obj <- AddMetaData(seurat_obj, cluster_results)
```

## Evaluate clustering results

For this section, we'll just use the SCE object.

### Silhouette width

Silhouette width values range from -1 to 1, where a value close to 1 means a given cell is very closely matched to its own cluster, and lower values mean a cell is not well matched to its own cluster.

We'll use the function `rOpenScPCA::calculate_silhouette` to calculate the silhouette width, and then visualize the results.
This function assumes columns called `cell_id` and `cluster`.

```{r}
silhouette_results <- calculate_silhouette(
  sce,
  cluster_results
)

silhouette_results
```

```{r}
ave_silhouette_width <- mean(silhouette_results$silhouette_width)
ggplot(silhouette_results) +
  aes(x = silhouette_width) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = ave_silhouette_width, color = "red", size = 1)
```

For most clusters, it is below 0.25, and for cluster 5 it is actually negative.
We also see it is higher for clusters 11 and 12, reaching up to about 0.5.

Overall, these values suggest that clustering is not fantastic and many cells may not well matched. 
We can explore this further by visualizing distributions for each cluster:

```{r}
ggplot(silhouette_results) +
  aes(x = cluster, y = silhouette_width) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  stat_summary(color = "red")
```
Cluster 5 looks particularly rough, but cluster 11 seems the most reliable.

### Neighborhood purity

We'll use the function `rOpenScPCA::calculate_purity` to calculate the neighborhood purity, and then visualize the results.
This function assumes columns called `cell_id` and `cluster`.

```{r}
purity_results <- calculate_purity(
  sce,
  cluster_results
)

purity_results
```

```{r}
ave_purity <- mean(purity_results$purity)

ggplot(purity_results) +
  aes(x = purity) +
  geom_density(fill = "lightblue") +
  geom_vline(xintercept = ave_purity, color = "red", size = 1)
```
Not bad! Let's look at individual clusters:

```{r}
ggplot(purity_results) +
  aes(x = cluster, y = purity) +
  geom_violin() +
  geom_jitter(size = 0.75, alpha = 0.25) +
  stat_summary(color = "red")
```
With this measure, clusters 1, 3, and 6 all have average values below 0.75, but the rest are high - in particular clusters 11, 12, and 13.

### Cluster stability

We'll use the function `rOpenScPCA::calculate_stability` to calculate the neighborhood purity, and then visualize the results.
This function assumes columns called `cell_id` and `cluster`.
